\vspace{-0.25em}\thetitle

### 2009-07-29 paper.tex
...

Before we can present our motivating observation about the linear structure of network traffic, we must explain how flow behaviors are represented as feature-frequency vectors. spaces.
The low-level details of this representation are given in \Section{representation}.
For now, what the reader must know is that each feature is represented by a vector where each possible value of the feature correspond to a dimension and the normalized frequency of that value occurring is the corresponding coordinate in the vector.
We have already given the example of packet size distribution in the introduction.
The vectors are normalized in that the sum of the coordinates of each feature-frequency vector is unity.
To combine multiple feature-frequency vectors into a composite description of the overall behavior of the flow, we take the direct sum their representations:\directsumnote
\begin{align}
  \text{flow} =
  \text{size} \directsum
  \text{ival} \directsum
  \text{type} \directsum
  \text{ports} \directsum
  \text{pkts}.
\end{align}
The behavior of a feature across a collection of flows can be represented as a matrix where each row represents a flow:
\begin{align}
  \text{Size} = \begin{bmatrix}
    \text{size}_1 \\
    \vdots \\
    \text{size}_m
  \end{bmatrix}.
\end{align}
The overall behavior of the collection of flows then becomes a concatenation of these feature matrices:
\begin{align}
  X = [ ~
  \text{Size} ~
  \text{Ival} ~
  \text{Type} ~
  \text{Ports} ~
  \text{Pkts} ~
  ].
\end{align}
It is this matrix, representing the total behavior of a collection of flows, to which we apply our analysis.

@@

We embed the following features of flow behavior into vector spaces via feature-frequency representation:
packet size distribution,
inter-packet interval distribution,
\caps{IP} protocol type,
\caps{TCP/UDP} port numbers,
and packet count.
Packet sizes are already naturally discrete, taking on only values from 1 to the \caps{MTU}, which was 1500 in all of our traces.
Inter-packet intervals must be quantized, using the scheme described by Karpinski~\emph{et~al.}~\cite{Karpinski08} but using only 500 dimensions instead of 1500.
This scheme maps bins of interval durations to indices, with the bins starting at the scale of milliseconds, and ramping smoothly up to seconds at the high end of the scale.
Intervals of more than ten minutes are considered to start a new flow.
\caps{IP} protocol type also naturally discrete, with values from 0 to 255, represented by a 256-dimensions histogram vector.
The feature-frequency vector for a single flow in this case is always just a vector with a single unit entry and all other entries zero.
Port numbers are naturally discrete as well, but the range of port numbers (0-65535) is too large for practical representation using a full histogram.
Instead, we represent port numbers by their remainder modulo 1000.
Port numbers are sufficiently spread out that this still provides ample distinction between traffic types.
Moreover, since the direction of a flow is generally uninteresting, we represent both source and destination in a single vector containing two values.
We also represent the packet count of each flow.
Since the potential packet count is unlimited, some approach is needed for limiting the dimension of this frequency representation.
We use the simplest approach and simply cap the packet count at 3000 packets:
all flows with 3000 or more packets fall into the same histogram bin and are simply considered ``very large.''

@@

>>>>

Before we can present our motivating observation about the linear structure of network traffic, we must explain how flow behaviors are represented as feature-frequency vectors. spaces.
The low-level details of this representation are given in \Section{representation}.
For now, what the reader must know is that each feature is represented by a vector where each possible value of the feature correspond to a dimension and the normalized frequency of that value occurring is the corresponding coordinate in the vector.
We have already given the example of packet size distribution in the introduction.
The vectors are normalized in that the sum of the coordinates of each feature-frequency vector is unity.
To combine multiple feature-frequency vectors into a composite description of the overall behavior of the flow, we take the direct sum their representations:\directsumnote
\begin{align}
  \text{flow} =
  \text{size} \directsum
  \text{ival} \directsum
  \text{type} \directsum
  \text{port} \directsum
  \text{pkts}.
\end{align}
The behavior of a feature across a collection of flows can be represented as a matrix where each row represents a flow:
\begin{align}
  \text{Size} = \begin{bmatrix}
    \text{size}_1 \\
    \vdots \\
    \text{size}_m
  \end{bmatrix}.
\end{align}
The overall behavior of the collection of flows then becomes a concatenation of these feature matrices:
\begin{align}
  X = [ ~
  \text{Size} ~
  \text{Ival} ~
  \text{Type} ~
  \text{Port} ~
  \text{Pkts} ~
  ].
\end{align}
It is this matrix, representing the total behavior of a collection of flows, to which we apply our analysis.

@@

We embed the following features of flow behavior into vector spaces via feature-frequency representation:
packet size distribution,
inter-packet interval distribution,
\caps{IP} protocol type,
\caps{TCP/UDP} port numbers,
and packet count.
Packet sizes are already naturally discrete, taking on only values from 1 to the \caps{MTU}, which was 1500 in all of our traces.
Inter-packet intervals must be quantized, using the scheme described by Karpinski~\emph{et~al.}~\cite{Karpinski08} but using only 500 dimensions instead of 1500.
This scheme maps bins of interval durations to indices, with the bins starting at the scale of milliseconds, but ramping smoothly up to seconds at the high end of the scale.
Intervals of more than ten minutes are considered to start a new flow.
\caps{IP} protocol type also naturally discrete, with values from 0 to 255, represented by a 256-dimensions histogram vector.
The feature-frequency vector for a single flow in this case is always just a vector with a single unit entry and all other entries zero.
Port numbers are naturally discrete as well, but the range of port numbers (0-65535) is too large for practical representation using a full histogram.
Instead, we represent port numbers by their remainder modulo 1000.
Port numbers are sufficiently spread out that this still provides ample distinction between traffic types.
Moreover, since the direction of a flow is generally uninteresting, we represent both source and destination in a single vector containing two values.
We also represent the packet count of each flow.
Since the potential packet count is unlimited, some approach is needed for limiting the dimension of this frequency representation.
We use the simplest approach and simply cap the packet count at 3000 packets:
all flows with 3000 or more packets fall into the same histogram bin and are simply considered ``very large.''

@@

\newfootnote{\svdnnnote}{The first singular vector of a nonnegative matrix is guaranteed to be nonnegative, but the next will generally not be.}

The next step towards recovering the structure of the hypothesized model is to determine the hull points of each linear substructure.
We have determined, as well as we can, which of the $q_i$ belong to the same structures.
Now we must determine the distributions $p_j$ which generate these structures.
We can use \caps{SVD} to determine the most prominent linear components of each structure, but \caps{SVD} will give us vectors of mixed sign.\svdnnnote
many invalid points will lie within the convex hull of these vectors.
\caps{SVD} can give orthogonal vectors spanning the structure, but we must turn to other techniques to recover possible vectors for the $p_j$.

In essence, what we have is a nonnegative matrix factorization (\caps{NMF}) problem:
if $Q_*$ is a sub-matrix of rows in the same structure class, we want to find nonnegative matrices, $W_*$ and $P_*$, such that $Q_* \approx W_* P_*$;
% We have the additional constraint that $W_*$ and $P_*$ have unit-sum rows, but that can easily be guaranteed since 
$P$ is then a vertical concatenation of small $P_*$ matrices, while $W$ is a direct sum of $W_*$ matrices.
Many \caps{NMF} algorithms have been proposed since Lee and Seung published the first~\cite{Lee01}.
We use Kim and Park's alternating non-negative least squares (\caps{ANLS}) algorithm~\cite{Kim08}.
It converges much more rapidly than the Lee and Seung algorithm, but stalls when it gets close to an optimum, so we refine that result by applying the Lee and Seung algorithm.

\newfootnote{\cddnote}{To do this, we use Avis and Fukuda's vertex enumeration algorithm as implemented in Fukuda's excellent CDD library~\cite{Avis92}.}

We find that using standard random \caps{NMF} initialization does not yield factorizations of sufficient quality, so we have developed the following initialization procedure.
Reduce the data to the number structural dimensions sought using \caps{SVD}.
Cluster the reduced data points into one cluster per structural dimension using $k$-means clustering.
Use cluster membership from this step to combine the original data, leaving only one composite row per desired structural dimension.
These rows provide a ``denoised'' set of nonnegative spanning vectors that should recover the true low-dimensional subspace from which the data were sampled.
Finally, we find the intersection of this subspace with the standard simplex in $\R^n$.
This step gives us a set of vertices geometrically guaranteed to contain all of the recovered low-dimensional subspace.\cddnote
